---
title: "Statistical Inference Course Project - Part 2"
author: "Jonathan Chang"
date: "February 28, 2017"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Overview

### 1. Exploratory Data Analysis
Load the ToothGrowth data and perform some basic exploratory data analyses

```{r}
library(datasets)
data(ToothGrowth)
head(ToothGrowth)
str(ToothGrowth)

ToothGrowth$dose <- as.factor(ToothGrowth$dose) # convert dose to factor
```

### 2. Basic Summary of the Data
Provide a basic summary of the data.
```{r}
library(ggplot2)
library(gridExtra)
summary(ToothGrowth)

# basic boxplot
supp_plot = ggplot(ToothGrowth, aes(supp, len, fill = supp)) + geom_boxplot()
dose_plot = ggplot(ToothGrowth, aes(dose, len, fill = dose)) + geom_boxplot()
grid.arrange(supp_plot, dose_plot, ncol = 2)
```

### 3. Compare Tooth Growth by supp and dose
Use confidence intervals and/or hypothesis tests to compare tooth growth by supp and dose. (Only use the techniques from class, even if there are other approaches worth considering)

We perform the following T tests.

By supplement type:
```{r}
t.test(len ~ supp, data=ToothGrowth)
```

By dose intervals:
```{r}
tooth_dose_half_one <- subset(ToothGrowth, dose %in% c(0.5, 1.0))
tooth_dose_half_two  <- subset(ToothGrowth, dose %in% c(0.5, 2.0))
tooth_dose_one_two <- subset(ToothGrowth, dose %in% c(1.0, 2.0))

t.test(len ~ dose, data=tooth_dose_half_one)
t.test(len ~ dose, data=tooth_dose_half_two)
t.test(len ~ dose, data=tooth_dose_one_two)
```

And by supplement type and dose level:
```{r}
tooth_dose_half <- subset(ToothGrowth, dose == 0.5)
tooth_dose_one <- subset(ToothGrowth, dose == 1.0)
tooth_dose_two <- subset(ToothGrowth, dose == 2.0)

t.test(len ~ supp, data=tooth_dose_half)
t.test(len ~ supp, data=tooth_dose_one)
t.test(len ~ supp, data=tooth_dose_two)
```

### 4. Conclusions
State your conclusions and the assumptions needed for your conclusions.

#### Assumptions

* the variables are independent and identically distributed
* tooth growth follows the normal distribution

#### Conclusions
For the tests where we obtained a p-value > 0.05 we could not reject the null hypothesis.
Based on the cases where the reported p-values were lower than 0.05, we conclude the following:

* Dosage is positively correlated with tooth growth (independent of supplement type).
* At lower dosages (.5 - 1 Mg), orange juice causes greater tooth growth than vitamin c.
